CFF GCT Study

### 2025-01-07 Christine Chan/CFF GCT Study/cff_gct_analysis.qmd
minor changes to GCT

---
title: "CFF GCT Study"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    fig-height: 9
    fig-width: 12
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: /home/tim/Documents/Miscellaneous/zotero.bib
csl: /home/tim/GitHub/styles/american-medical-association.csl
editor: source
---

@@

```{r data cleaning}
#| warning: false
# Import from REDCap
df <- read.csv("./Christine Chan/CFF GCT Study/Data_Raw/CysticFibrosisScreen_DATA_LABELS_2024-04-19_1740.csv",
  na.strings = ""
)
# Create standardized study ID
df$ID <- sub("--1|--2|--01|-1|-2", "", df[, 1])
# Numeric variables
df$Plasma.Glucose <- as.numeric(df$Plasma.Glucose)
df$X1.hour.Plasma.Glucose.Value <- as.numeric(df$X1.hour.Plasma.Glucose.Value)
# Combine race and ethnicity columns
races <- c(
  "White", "Black/African American", "American Indian/Alaska Native",
  "Asian Non-Indian", "Asian Indian", "Native Hawaiian/Other Pacific Islander"
)
ethnicities <- c("Hispanic/Latino", "Not Hispanic/Latino", "N/A")
df$Race <- apply(
  df[, grep("Ethnic.Origin.Race", colnames(df))], 1,
  function(r) {
    paste0(races[which(r == "Checked")], collapse = "/")
  }
)
df$Race <- factor(df$Race, levels = races)
df$Ethnicity <- apply(
  df[, grep("Ethnicity", colnames(df))], 1,
  function(r) {
    paste0(ethnicities[which(r == "Checked")], collapse = "/")
  }
)
df$Ethnicity <- factor(df$Ethnicity, levels = ethnicities[1:2])
# Fill in missing data for every column and take the first row, then select and
# rename columns
df <- df %>%
  group_by(ID) %>%
  fill(everything(), .direction = "downup") %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  rename(
    "OGTT 0 min" = Plasma.Glucose..pre.glucose.drink...mg.dl.,
    "OGTT 60 min" = Plasma.Glucose..1.hour.after.glucose.drink...mg.dl.,
    "OGTT 120 min" = Plasma.Glucose..2.hour.after.glucose.drink...mg.dl.,
    "GCT 0 min plasma" = Plasma.Glucose,
    "GCT 0 min capillary" = Capillary,
    "GCT 60 min plasma" = X1.hour.Plasma.Glucose.Value,
    "GCT 60 min capillary" = X1.hour.Capillary.Glucose.Value,
    "HbA1c" = Hemoglobin.A1c..,
    "Age" = Age.at.enrollment..years.,
    "Pancreatic insufficient" = Pancreatic.exocrine.insufficient
  ) %>%
  select(
    ID, HbA1c, Age, Gender, Race, Ethnicity, `Pancreatic insufficient`,
    contains("GCT"), contains("OGTT ")
  )
# Get diagnosis
df$Diagnosis <- "NGT"
df$Diagnosis[df$`OGTT 0 min` >= 100] <- "IFG"
df$Diagnosis[df$`OGTT 60 min` >= 200] <- "INDET"
df$Diagnosis[df$`OGTT 120 min` >= 140] <- "IGT"
df$Diagnosis[df$`OGTT 120 min` >= 200] <- "CFRD"
df[which(rowSums(is.na(df[, grep("OGTT", colnames(df))])) == 3), "Diagnosis"] <-
  NA
df$Diagnosis <- factor(df$Diagnosis,
  levels = c("NGT", "IFG", "INDET", "IGT", "CFRD")
)
df$Dysglycemia <- factor(df$Diagnosis,
  levels = c("NGT", "IFG", "INDET", "IGT", "CFRD"),
  labels = c("NGT", "AGT", "AGT", "AGT", "AGT")
)
df$CFRD <- factor(df$`OGTT 120 min` >= 200,
  levels = c(F, T),
  labels = c("No", "Yes")
)
# Write for Christine to check
write.csv(df,
  file = "./Christine Chan/CFF GCT Study/Data_Clean/gct_analysis_data.csv",
  row.names = F, na = ""
)
```

@@

```{r}
#| label: table1a
#| tbl-cap: Participant characteristics (including those missing OGTT data)
#| message: false
df %>%
  select(-ID, -Dysglycemia) %>%
  tbl_summary(missing_text = "(Missing)")
# Now remove those without OGTT
df <- df %>% filter(!is.na(Diagnosis))
```

```{r}
#| label: table1b
#| tbl-cap: Participant characteristics by diagnosis
#| message: false
df %>%
  select(-ID, -Dysglycemia) %>%
  tbl_summary(by = Diagnosis, missing_text = "(Missing)") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

```{r}
#| label: fig-boxplot
#| fig-cap: Boxplot of OGTT glucose values by diabetes status
#| warning: false
plot_df <- df %>% pivot_longer(contains("OGTT"),
  names_to = "OGTT Timepoint", values_to = "Glucose"
)
plot_df$`OGTT Timepoint` <- factor(plot_df$`OGTT Timepoint`,
  levels = c("OGTT 0 min", "OGTT 60 min", "OGTT 120 min")
)
box_p <- ggplot(
  plot_df,
  aes(x = `OGTT Timepoint`, y = Glucose, fill = `OGTT Timepoint`)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = c(100, 140, 200, 300)) +
  facet_wrap(~Diagnosis) +
  theme_bw() +
  theme(legend.position = "none")
box_p
```

```{r}
#| label: fig-dot
#| fig-cap: Dot plot of OGTT glucose values by diabetes status
#| warning: false
dot_p <- ggplot(
  plot_df,
  aes(x = `OGTT Timepoint`, y = Glucose, color = `OGTT Timepoint`)
) +
  geom_jitter() +
  scale_y_continuous(breaks = c(100, 140, 200, 300)) +
  facet_wrap(~Diagnosis) +
  theme_bw() +
  theme(legend.position = "none")
dot_p
```

```{r}
#| label: fig-violin
#| fig-cap: Violin plot of OGTT glucose values by diabetes status
#| warning: false
violin_p <- ggplot(
  plot_df,
  aes(x = `OGTT Timepoint`, y = Glucose, fill = `OGTT Timepoint`)
) +
  geom_violin() +
  scale_y_continuous(breaks = c(100, 140, 200, 300)) +
  facet_wrap(~Diagnosis) +
  theme_bw() +
  theme(legend.position = "none")
violin_p
```

>>>>

---
title: "CFF GCT Study"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    fig-height: 9
    fig-width: 12
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: /Users/timvigers/Library/CloudStorage/Dropbox/Miscellaneous/zotero.bib
csl: /Users/timvigers/Library/CloudStorage/Dropbox/GitHub/styles/american-medical-association.csl
editor: source
---

@@

```{r data cleaning}
#| warning: false
# Import from REDCap
df <- read.csv("./Christine Chan/CFF GCT Study/Data_Raw/CysticFibrosisScreen_DATA_LABELS_2024-04-19_1740.csv",
  na.strings = ""
)
# Create standardized study ID
df$ID <- sub("--1|--2|--01|-1|-2", "", df[, 1])
# Numeric variables
df$Plasma.Glucose <- as.numeric(df$Plasma.Glucose)
df$X1.hour.Plasma.Glucose.Value <- as.numeric(df$X1.hour.Plasma.Glucose.Value)
# Combine race and ethnicity columns
races <- c(
  "White", "Black/African American", "American Indian/Alaska Native",
  "Asian Non-Indian", "Asian Indian", "Native Hawaiian/Other Pacific Islander"
)
ethnicities <- c("Hispanic/Latino", "Not Hispanic/Latino", "N/A")
df$Race <- apply(
  df[, grep("Ethnic.Origin.Race", colnames(df))], 1,
  function(r) {
    paste0(races[which(r == "Checked")], collapse = "/")
  }
)
df$Race <- factor(df$Race, levels = races)
df$Ethnicity <- apply(
  df[, grep("Ethnicity", colnames(df))], 1,
  function(r) {
    paste0(ethnicities[which(r == "Checked")], collapse = "/")
  }
)
df$Ethnicity <- factor(df$Ethnicity, levels = ethnicities[1:2])
# Fill in missing data for every column and take the first row, then select and
# rename columns
df <- df %>%
  group_by(ID) %>%
  fill(everything(), .direction = "downup") %>%
  filter(row_number() == 1) %>%
  ungroup() %>%
  rename(
    "OGTT 0 min" = Plasma.Glucose..pre.glucose.drink...mg.dl.,
    "OGTT 60 min" = Plasma.Glucose..1.hour.after.glucose.drink...mg.dl.,
    "OGTT 120 min" = Plasma.Glucose..2.hour.after.glucose.drink...mg.dl.,
    "GCT 0 min plasma" = Plasma.Glucose,
    "GCT 0 min capillary" = Capillary,
    "GCT 60 min plasma" = X1.hour.Plasma.Glucose.Value,
    "GCT 60 min capillary" = X1.hour.Capillary.Glucose.Value,
    "HbA1c" = Hemoglobin.A1c..,
    "Age" = Age.at.enrollment..years.,
    "Pancreatic insufficient" = Pancreatic.exocrine.insufficient
  ) %>%
  select(
    ID, HbA1c, Age, Gender, Race, Ethnicity, `Pancreatic insufficient`,
    contains("GCT"), contains("OGTT ")
  )
# Per Christine, we can exclude C046's 140 value and keep them in the NGT
# category because they had multiple normal OGTTs after this one and the 140 is
# likely a transcription error
df$`OGTT 0 min`[df$ID == "C046"] <- NA
# Get diagnosis
df$Diagnosis <- "NGT"
df$Diagnosis[df$`OGTT 0 min` >= 100] <- "IFG"
df$Diagnosis[df$`OGTT 60 min` >= 200] <- "INDET"
df$Diagnosis[df$`OGTT 120 min` >= 140] <- "IGT"
df$Diagnosis[df$`OGTT 120 min` >= 200] <- "CFRD"
df[which(rowSums(is.na(df[, grep("OGTT", colnames(df))])) == 3), "Diagnosis"] <-
  NA
df$Diagnosis <- factor(df$Diagnosis,
  levels = c("NGT", "IFG", "INDET", "IGT", "CFRD")
)
df$Dysglycemia <- factor(df$Diagnosis,
  levels = c("NGT", "IFG", "INDET", "IGT", "CFRD"),
  labels = c("NGT", "AGT", "AGT", "AGT", "AGT")
)
df$Diabetes <- factor(df$Diagnosis,
  levels = c("NGT", "IFG", "INDET", "IGT", "CFRD"),
  labels = c("NGT", "Pre-CFRD", "Pre-CFRD", "Pre-CFRD", "CFRD")
)
df$CFRD <- factor(df$`OGTT 120 min` >= 200,
  levels = c(F, T),
  labels = c("No", "Yes")
)
# Has OGTT vs. does not
df$`Got OGTT and GCT?` <- apply(
  df[, c("OGTT 120 min", "GCT 60 min plasma")], 1, function(r) {
    !any(is.na(r))
  }
)
df$`Got OGTT and GCT?` <- factor(df$`Got OGTT and GCT?`,
  levels = c(T, F),
  labels = c(
    "Have GCT 1 Hour Plasma or OGTT 2 Hour",
    "Missing GCT 1 Hour Plasma or OGTT 2 Hour"
  )
)
# Write for Christine to check
write.csv(df,
  file = "./Christine Chan/CFF GCT Study/Data_Clean/gct_analysis_data.csv",
  row.names = F, na = ""
)
```

@@

```{r}
#| label: table1a
#| tbl-cap: Participant characteristics by OGTT status
#| message: false
df %>%
  select(-ID) %>%
  tbl_summary(by = `Got OGTT and GCT?`, missing_text = "(Missing)") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
# Limit to 189 after this
df <- df %>%
  filter(`Got OGTT and GCT?` == "Have GCT 1 Hour Plasma or OGTT 2 Hour") %>%
  select(-`Got OGTT and GCT?`)
```

```{r}
#| label: table1b
#| tbl-cap: Participant characteristics by CFRD status
#| message: false
df %>%
  select(-ID) %>%
  tbl_summary(by = Diabetes, missing_text = "(Missing)") %>%
  add_overall() %>%
  add_p() %>%
  separate_p_footnotes()
```

@@

```{r}
#| label: fig-boxplot
#| fig-cap: Boxplot of OGTT glucose values by diabetes status
#| warning: false
plot_df <- df %>%
  pivot_longer(contains("OGTT"),
    names_to = "OGTT Timepoint", values_to = "Glucose"
  )
plot_df$`OGTT Timepoint` <- factor(plot_df$`OGTT Timepoint`,
  levels = c("OGTT 0 min", "OGTT 60 min", "OGTT 120 min")
)
box_p <- ggplot(
  plot_df,
  aes(x = `OGTT Timepoint`, y = Glucose, fill = `OGTT Timepoint`)
) +
  geom_boxplot() +
  scale_y_continuous(breaks = c(100, 140, 200, 300)) +
  facet_wrap(~Diabetes) +
  theme_bw(base_size = 20) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title.x = element_blank()
  )
box_p
```
